### 2015-03-21 background.tex
initial text for bug finding section
many updates to refocus in drivers

Nowadays, multicore architectures have become mainstream, leading to an ever-increasing demand for concurrency in operating systems. Modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same module concurrently; a running module process can block and be replaced by another process in the same module; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging). All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:

@@

Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux module. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.

\begin{lstlisting}[caption = Simple data race in a Linux module, label = fig:data_race_example]
static void ep1 (struct shared *dev) {
  dev->resource = 5;
}

@@

Lockset analysis avoids reasoning about arbitrary thread interleavings, and thus has the potential to scale well in realistic concurrent programs. The technique, though, suffers from imprecision (i.e. can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, as a classic dynamic analyser, Eraser's bug finding ability is limited to the execution paths that the tool explores. To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.

>>>>

Nowadays, multicore architectures have become mainstream, leading to an ever-increasing demand for concurrency in operating systems. Modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging). All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:

@@

Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux driver. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.

\begin{lstlisting}[caption = Simple data race in a Linux driver, label = fig:data_race_example]
static void ep1 (struct shared *dev) {
  dev->resource = 5;
}

@@

Lockset analysis avoids reasoning about arbitrary thread interleavings, and thus has the potential to scale well in realistic concurrent programs. The technique, though, suffers from imprecision (i.e. can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, as a classic dynamic analyser, Eraser's bug-finding ability is limited to the execution paths that the tool explores. To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.

### 2015-03-21 conclusion.tex
many updates to refocus in drivers

In this paper, we presented \whoop, a novel infrastructure for (i) analysing the concurrent behaviour of Linux kernel modules and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis. In the heart of our approach, we combine static lock set analysis with sound sequentialisation and automated theorem proving. Our lock set analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.

>>>>

In this paper, we presented \whoop, a novel infrastructure for (i) analysing the concurrent behaviour of Linux drivers and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis. In the heart of our approach, we combine static lock set analysis with sound sequentialisation and automated theorem proving. Our lock set analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.

### 2015-03-21 intro.tex
many updates to refocus in drivers

Linux kernel modules, such as device drivers and filesystems, exist at the thin boundary between hardware and software. They are complex pieces of system-level software responsible for the interaction between a user application, the operating system and any hardware devices that are attached to a computer. Linux kernel modules are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}. Especially in the case of drivers, this has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.

@@

A major reason behind these concurrency bugs is that Linux kernel modules operate inside a complicated and hostile environment~\cite{corbet2005linux}. For example the Linux kernel is able to invoke its modules from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation. All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms (such as locks) are in place to control concurrent access to shared resources.

@@

In this paper, we present \whoop, a novel infrastructure for (i) automatically analysing the concurrent behaviour of Linux kernel modules written in the C language and (ii) accelerating external bug-finders that can easily plug in our toolchain. \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux kernel modules (e.g. device drivers and filesystems). The key idea behind our approach is that a module can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyse device drivers and other concurrent programs. Using \whoop and \corral we analysed 10 modules available in the Linux distribution.

To summarize, our contributions are as follows:
\begin{itemize}
\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux kernel modules and accelerating a plugged-in precise bug-finder.
\item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux kernel modules.
\item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
\item We analyse 10 Linux kernel modules, including various drivers and a filesystem.
\end{itemize}

>>>>

Device drivers exist at the thin boundary between hardware and software. They are complex pieces of system-level software responsible for the interaction between the operating system and any hardware devices that are attached to a computer. Drivers are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}. This has a negative impact in hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.

@@

A major reason behind these concurrency bugs is that device drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}. For example the Linux kernel is able to invoke its drivers from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation. All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms (such as locks) are in place to control concurrent access to shared resources.

@@

In this paper, we present \whoop, a novel infrastructure for (i) automatically analysing the concurrent behaviour of Linux drivers written in the C language and (ii) accelerating external bug-finders that can easily plug in our toolchain. \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers (e.g. device drivers and filesystems). The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyse device drivers and other concurrent programs. Using \whoop and \corral we analysed 10 drivers available in the Linux distribution.

To summarize, our contributions are as follows:
\begin{itemize}
\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux drivers and accelerating a plugged-in precise bug-finder.
\item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux drivers.
\item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
\item We analyse 10 Linux drivers, including various drivers and a filesystem.
\end{itemize}

### 2015-03-21 staticanalysis.tex
many updates to refocus in drivers

After the original program has been translated to the abstract Boogie program, the core \whoop tool is invoked, which performs \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.

@@

We now formalise our approach to verifying data race freedom in a kernel module by statically analysing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a module, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this module. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all modules that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking. The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.

@@

\begin{theorem}
\label{theorem:locksets}
For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the kernel module with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
\end{theorem}

@@

If \whoop cannot infer a lock it will exit with a warning. A limitation of lockset analysis is that it cannot understand external locking mechanisms (\whoop currently only supports Linux kernel mutexes and spinlocks). It is relatively straightforward to enhance \whoop, though, with new locking primitives. \whoop will also exit with a warning if it detects ``improper'' use of locks (e.g. dynamic lock creation). The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practises when developing modules for the Linux kernel.

@@

Early on during the development of \whoop, we realised that scalability would be a serious issue. When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code) with all functions fully inlined, we were running out of memory. One reason was that the C to Boogie translation produced hundreds of thousands of Boogie IVL code, which could not fit in memory. Another reason was that some entry points recursively call helper functions. Inherently, recursion does not work with inlining. To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux kernel modules in a scalable fashion, but without destroying precision.

@@

To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel uses its various modules. This is an ongoing manual effort: the more kernel modules we study, the more domain-specific properties we discover, which we then use for increased precision.

>>>>

After the original program has been translated to the abstract Boogie program, the core \whoop tool is invoked, which performs \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux drivers. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.

@@

We now formalise our approach to verifying data race freedom in a driver by statically analysing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking. The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.

@@

\begin{theorem}
\label{theorem:locksets}
For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the driver with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
\end{theorem}

@@

If \whoop cannot infer a lock it will exit with a warning. A limitation of lockset analysis is that it cannot understand external locking mechanisms (\whoop currently only supports Linux kernel mutexes and spinlocks). It is relatively straightforward to enhance \whoop, though, with new locking primitives. \whoop will also exit with a warning if it detects ``improper'' use of locks (e.g. dynamic lock creation). The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practises when developing drivers for the Linux kernel.

@@

Early on during the development of \whoop, we realised that scalability would be a serious issue. When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code) with all functions fully inlined, we were running out of memory. One reason was that the C to Boogie translation produced hundreds of thousands of Boogie IVL code, which could not fit in memory. Another reason was that some entry points recursively call helper functions. Inherently, recursion does not work with inlining. To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux drivers in a scalable fashion, but without destroying precision.

@@

To increase precision we enhance \whoop with domain-specific knowledge of how the Linux kernel uses its various drivers. This is an ongoing manual effort: the more drivers we study, the more domain-specific properties we discover, which we then use for increased precision.

### 2015-03-21 translation.tex
fix
many updates to refocus in drivers

The main aim of this phase is to take an input Linux kernel module written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g. Boogie and \corral). To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.

First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given module, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points. This function acts as the main entry point to the module after the latter has been registered with the kernel. This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{staticanalysis} and~\ref{bugfinding}). The output is a file that contains information regarding the entry points and a file with the rewritten program.

@@

Example~\ref{fig:original_program} shows a simple program in C and Example~\ref{fig:smack_translation} shows its corresponding translation in Boogie IVL using SMACK. The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource. Notice that functional calls (e.g. to lock and unlock a mutex) are preserved in the translation.

>>>>

The main aim of this phase is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g. Boogie and \corral). To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.

First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given driver, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points. This function acts as the main entry point to the driver after the latter has been registered with the kernel. This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{staticanalysis} and~\ref{bugfinding}). The output is a file that contains information regarding the entry points and a file with the rewritten program.

@@

Example~\ref{fig:original_program} shows a simple program in C and Example~\ref{fig:smack_translation} shows its corresponding translation in Boogie IVL using SMACK. The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource. Notice that function calls (e.g. to lock and unlock a mutex) are preserved in the translation.

### 2015-03-21 whoop.tex
many updates to refocus in drivers

We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in bug-finder.

@@

Figure~\ref{fig:whoop} depicts the \whoop infrastructure. Initially, \whoop accepts a kernel module written in C, together with a Linux environmental model. The environmental model (i.e. stub Linux header files) is required to ``close'' the module and allow it to be translated and subsequently analysed. Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs its static lockset analysis (see Section~\ref{staticanalysis}) to detect all potential data races in the abstract program. Finally, \whoop uses information from the static analysis phase to accelerate bug-finding with \corral (see Section~\ref{bugfinding}).

>>>>

We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux drivers for data races and (ii) accelerate a plugged-in bug-finder.

@@

Figure~\ref{fig:whoop} depicts the \whoop infrastructure. Initially, \whoop accepts a driver written in C, together with a Linux environmental model. The environmental model (i.e. stub Linux header files) is required to ``close'' the driver and allow it to be translated and subsequently analysed. Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs its static lockset analysis (see Section~\ref{staticanalysis}) to detect all potential data races in the abstract program. Finally, \whoop uses information from the static analysis phase to accelerate bug-finding with \corral (see Section~\ref{bugfinding}).
